Language learning in context via recursive pragmatic reasoning

### 2013-05-31 praglearn.tex
tweak

Formally, let an unadorned $L$ and $S$ denote our convention-based
listener and speaker. By assumption (c), the top-level listener has uncertainty over the lexicon; if the lexicon were known they would draw inferences as in $L_n$ above:
%\begin{align*}
%  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
%    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior. \\
%    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
%  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
%\end{align*}
%
\begin{align*}
  P_L(\obj | \word, \text{$L$'s data}) &=  \sum_\lex P_{L_n}(\obj | \word, \lex) P(\lex | \text{$L$'s data}) \\
\end{align*}
Here $L$'s data consists of $S$'s utterances together with whatever
other information $L$ may have about their meaning (e.g.~from
contextual clues). 
By assumption (b), $L$ treats these  utterances
as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus
as being informative about the lexicon. For instance, when the data is a set of fully observed 
word-referent pairs $\{w_i, o_i\}$:
\begin{align*}
  P(\lex | \text{$L$'s data}) \propto P(\lex) \prod_i P_{S_{n-1}}(w_i|o_i,  \lex)
\end{align*}
The top-level speaker attempts to select the word which soft-maximizes
their utility, with utility now being defined in terms of the
informativity of the expectation (over lexicons) that the listener
will have for the right referent\footnote{An alternative model would
  have the speaker take the expectation over informativity, instead of
  the informativity of the expectation.}:
\begin{align*}
  P_S(\word | \obj, \text{$S$'s data}) &\propto \exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
\end{align*}
Here $P(\lex|\text{$S$'s data})$ is defined similarly, when $S$ observes
$L$'s interpretations of various utterances, and treats them as
samples from $L_n$, not $L$.

@@

The recursive pragmatic reasoning portions of our model capture
findings on specificity implicature. Consider a lexicon in which the
rows represent the meaning of ``some'' and ``all'', and the columns
indicate worlds where \textsc{some-but-not-all} and \textsc{all} are
true, respectively: $\left[
    \begin{array}{cc}
      0.5 & 0.5 \\
      0 & 1
    \end{array}
  \right]$. This lexicon captures the regularity that ``some'' can be said of
  situations where {\sc all} is actually the case. Assuming uniform
  costs for the two words, a rational speaker confronted with such a
  lexicon will prefer to say ``some'' only when they mean \textsc{some
    but not all}, and this is therefore the interpretation that a
  rational listener -- and our model -- will place on ``some'' as well.

@@

@@

Experimental results suggest that communicators are able to establish
novel, consensus-based communication systems. For example, adults
playing a communication game using only novel symbols with no
conventional meaning will typically converge on a set of new
conventions which allow them to accomplish their task
\cite{galantucci2005}. Or in a less extreme example, communicators
asked to refer to novel objects invent conventional names for them
over the course of repeated interactions (e.g., ``the ice skater'' for
an abstract figure vaguely resembling an ice skater,
\cite{clark1986}). From a pure learning perspective this behavior is anomalous,
however: Since both agents know perfectly well that there is no
existing convention to discover, there is nothing to learn from the
other's behavior. Furthermore, even if only one partner is producing
the novel expressions, their behavior still becomes more regular
(conventional) over time, which would seem to rule out a role for
learning---even if there is some pattern in the expressions the
speaker chooses to use, there is certainly nothing for the
\textit{speaker} to learn by observing these patterns, and thus their
behavior should not change over time.
% An obvious hack might be to simply stipulate that speakers take
% their own utterances as training data for further learning, but this
% would fail to explain yet another wrinkle in this phenomenon:
% speakers only regularize their production over time if they have
% some kind of feedback indicating that the listener understands; if
% such feedback is removed, the phenomenon is dramatically reduced (X X
% krauss 1966, branigan catchpole pickering 2011).

@@

A stronger example of how pragmatics can create communicative biases can be
observed by considering the Horn implicatures introduced in \S\ref{sec:horn-implicature}. 
In these games, pragmatic reasoning causes each agent to start out weakly biasedtowards the associations ``cheap'' $\leftrightarrow$
\textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}; this bias arises because their prior on literal meaning
is uniform and uninformative. A fully rational listener, therefore,
observing an uncertain speaker using this mapping, would discount it
as arising from this bias, and conclude that the speaker in fact did
not know which convention was in use. Conventional listeners,
however, believe that speakers do know which convention is in
use, and therefore tend to take their behavior as positive evidence that the
``good'' system is in use. Similarly, conventional speakers
will wager that since on average they will succed more often if
listeners are using the ``good'' system, they might as well try
it. When they succeed, they take their success as evidence that the listener was in
fact using the good system all along. As a result, dyads in this game
 end up converging onto a stable system at a rate far above
chance (Fig.~\ref{fig:horn}).

>>>>

Formally, let an unadorned $L$ and $S$ denote our convention-based
listener and speaker. By assumption (c), the top-level listener has uncertainty over the lexicon; if the lexicon were known they would draw inferences as in $L_n$ above:
%\begin{align*}
%  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
%    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior. \\
%    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
%  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
%\end{align*}
%
\begin{align*}
  P_L(\obj | \word, \text{$L$'s data}) &=  \sum_\lex P_{L_n}(\obj | \word, \lex) P(\lex | \text{$L$'s data}) \\
\end{align*}
Here $L$'s data consists of $S$'s utterances together with whatever
other information $L$ may have about their meaning (e.g.~from
contextual clues). 
By assumption (b), $L$ treats these  utterances
as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus
as being informative about the lexicon. For instance, when the data is a set of fully observed 
word-referent pairs $\{w_i, o_i\}$:
\begin{align*}
  P(\lex | \text{$L$'s data}) \propto P(\lex) \prod_i P_{S_{n-1}}(w_i|o_i,  \lex)
\end{align*}
The top-level speaker attempts to select the word which soft-maximizes
their utility, with utility now being defined in terms of the
informativity of the expectation (over lexicons) that the listener
will have for the right referent\footnote{An alternative model would
  have the speaker take the expectation over informativity, instead of
  the informativity of the expectation.}:
\begin{align*}
  P_S(\word & | \obj, \text{$S$'s data})  \propto \\
 &\exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
\end{align*}
Here $P(\lex|\text{$S$'s data})$ is defined similarly, when $S$ observes
$L$'s interpretations of various utterances, and treats them as
samples from $L_n$, not $L$.

@@

The recursive pragmatic reasoning portions of our model capture
findings on specificity implicature. Consider a lexicon in which the
rows represent the meaning of ``some'' and ``all'', and the columns
indicate worlds where \textsc{some-but-not-all} and \textsc{all} are
true, respectively: $\left[
    \begin{array}{cc}
      0.5 & 0.5 \\
      0 & 1
    \end{array}
  \right]$. This lexicon captures the regularity that ``some'' can be said of
  situations where {\sc all} is actually the case. Assuming uniform
  costs for the two words, a rational speaker confronted with such a
  lexicon will prefer to say ``some'' only when they mean \textsc{some
    but not all}, and this is therefore the interpretation that a
  rational listener---and our model---will place on ``some'' as well.

@@

\vspace{-2ex}

@@

Experimental results suggest that communicators are able to establish
novel, consensus-based communication systems. For example, adults
playing a communication game using only novel symbols with no
conventional meaning will typically converge on a set of new
conventions which allow them to accomplish their task
\cite{galantucci2005}. Or in a less extreme example, communicators
asked to refer to novel objects invent conventional names for them
over the course of repeated interactions (e.g., ``the ice skater'' for
an abstract figure vaguely resembling an ice skater,
\cite{clark1986}). From a pure learning perspective this behavior is anomalous,
however: Since both agents know perfectly well that there is no
underlying convention to discover, there is nothing to learn from the
other's behavior. Furthermore, even if only one partner is producing
the novel expressions, their behavior still becomes more regular
(conventional) over time, which would seem to rule out a role for
learning---even if there is some pattern in the expressions the
speaker chooses to use, there is certainly nothing for the
\textit{speaker} to learn by observing these patterns, and thus their
behavior should not change over time.
% An obvious hack might be to simply stipulate that speakers take
% their own utterances as training data for further learning, but this
% would fail to explain yet another wrinkle in this phenomenon:
% speakers only regularize their production over time if they have
% some kind of feedback indicating that the listener understands; if
% such feedback is removed, the phenomenon is dramatically reduced (X X
% krauss 1966, branigan catchpole pickering 2011).

@@

A stronger example of how pragmatics can create communicative biases can be
observed by considering the Horn implicatures introduced in \S\ref{sec:horn-implicature}. 
In these games, pragmatic reasoning causes each agent to start out weakly biasedtowards the associations ``cheap'' $\leftrightarrow$
\textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}; this bias arises because their prior on literal meaning
is uniform and uninformative. A fully rational listener, therefore,
observing an uncertain speaker using this mapping, would discount it
as arising from this bias, and conclude that the speaker in fact did
not know which convention was in use. Conventional listeners,
however, believe that speakers do know which convention is in
use, and therefore tend to take their behavior as positive evidence that the
``good'' system is in use. Similarly, conventional speakers
will wager that since on average they will succed more often if
listeners are using the ``good'' system, they might as well try
it. When they succeed, they take their success as evidence that the listener was in
fact using the good system all along. As a result, dyads in this game
not only end up converging onto a stable system at a rate far above
chance (Fig.~\ref{fig:horn}.
